Towards a general mathematical theory of experimental science

### 2019-01-28 Articles/TowardsAGeneralMathematicalTheoryOfExperimentalScience/article.tex
Adapting to FoP style
Updating accepted reference
Rewrote the introduction

\documentclass[letterpaper]{article}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

%% Sets page size and margins
\usepackage[letterpaper,top=1in,bottom=1in,left=1in,right=1in,marginparwidth=0.5in]{geometry}

% Temporary to remove "submitted to"
\makeatletter
\def\ps@pprintTitle{%
	\let\@oddhead\@empty
	\let\@evenhead\@empty
	\def\@oddfoot{}%
	\let\@evenfoot\@oddfoot}
\makeatother

\usepackage{lineno,hyperref}

\usepackage{amsmath, amsthm, amsfonts}
\usepackage[only,llbracket, rrbracket,llparenthesis,rrparenthesis]{stmaryrd}

@@

@@

% Allows to create negation symbols
\usepackage{MnSymbol}

@@

\title{Towards a general mathematical theory of experimental science}
\author{Gabriele Carcassi, Christine A. Aidala \\ University of Michigan}

\date{\today}

@@

\begin{abstract}
	In this article we lay the groundwork for a general mathematical theory of experimental science. The starting point will be the notion of verifiable statements, those assertions that can be shown to be true with an experimental test. We study the algebra of such objects and show how it is closed only under finite conjunction and countable disjunction. With simple constructions, we show that the set of possible cases distinguishable by verifiable statements is equipped with a natural Kolmogorov and second countable topology and a natural $\sigma$-algebra. This gives a clear physical meaning to those mathematical structures and provides a strong justification for their use in science. It is our hope and belief that such an approach can be extended to many areas of fundamental physics and will provide a consistent vocabulary across scientific domains.
\end{abstract}

@@

During the course of the last century a few fields of science and engineering have developed an associated general theory. For example, in computer science theory one starts by defining what a computational device is~\cite{Turing} and then shows that no algorithm exists that can correctly decide whether an arbitrary program terminates given an arbitrary input~\cite{Sipser}. In communications one defines what information is~\cite{Shannon} and shows that no encoding can outperform the Huffman coding~\cite{Pierce}. In control theory one defines what a system is~\cite{Brogan} and then looks for general strategies for control such as Kalman filtering~\cite{Kalman}. These theories are general in the sense that their results respectively apply to all control systems, all communication systems and all computational devices. That is, the results apply because of the mere definition of the subject matter.

No such general theory currently exists for physics, which is essentially composed of different theories loosely connected to each other, each with their own realm of applicability. However, we believe such a general theory can be created and in this work we present our basic strategy towards it. As one familiar with the general theories of other subjects would expect, we only need to focus on clarifying exactly what our starting points are and let the logic follow. That is, once we formalize exactly what it means to study a system through experimentation, we can reach interesting results that necessarily apply to all scientific theories. For example, we will see that any set of objects that are physically distinguishable must have cardinality no greater than continuum. We will see that defining what can be experimentally verified about them will always induce a topology that is Kolmogorov and second countable. These results are general because they apply independently of what is being studied.

One should not confuse this approach with the idea of a grand-unified theory or theory of everything. In that program, one looks for the theory of fundamental constituents from which all other theories can be derived. Our program is altogether different and orthogonal. Our subject matter is the scientific models themselves. Our aim is to understand what aspects of a scientific theory are already constrained by the mere fact that it is a scientific theory or by the assumptions one builds into them. While this general theory may be of some utility to someone working on a grand-unified theory as it may help narrow the search, neither one will lead to the other.

In this paper we briefly present the starting point for our general mathematical theory of experimental science. It is the product of a decade spent in reverse engineering the laws of particle mechanics~\cite{Carc1} and our subsequent work on formalizing the starting points~\cite{Carc2}. Our overall program is large and ambitious and a journal article cannot treat it comprehensively. Therefore we will only present an overview of the basic concepts, which are already well consolidated and provide a good template to understand our approach. We will only provide a sketch for the proofs, which can be found in the draft of our open access book~\cite{Carc3} together with more details, discussions and insights.

The basic idea is that science deals with verifiable statements: assertions whose truth can be tested experimentally. We study the logic of these statements, which is different from the standard Boolean logic. We then group them into experimental domains, which represent a set of verifiable statements that can be tested in an indefinite amount of time. From these we construct theoretical domains, which represent all possible predictions, and within these we find the possibilities, those predictions that determine the outcome of all possible tests. We will find that possibilities have a natural topology induced by the experimental domain and a natural $\sigma$-algebra induced by the theoretical domain. We will find that the topology is Kolmogorov and second countable, which limits the cardinality of the set of possibilities of a domain to the one of continuum. Note that these structures are indeed pervasive in science, and form the foundation for differential geometry (Riemannian and symplectic), measure theory, probability theory and many other mathematical tools already used in physics and other sciences.

@@

	\bibitem{Carc2} C. A. Aidala, G. Carcassi, M. J. Greenfield: Topology and experimental distinguishability. arXiv:1708.05492, 2017.

>>>>

\newif\ifarchive
\archivetrue
%\archivefalse

\ifarchive
	\documentclass[letterpaper]{article}
\else
	\documentclass[smallextended]{svjour3}
\fi

%% Language and font encodings
%\usepackage[english]{babel}
%\usepackage[utf8x]{inputenc}
%\usepackage[T1]{fontenc}

%% Sets page size and margins
%\usepackage[letterpaper,top=1in,bottom=1in,left=1in,right=1in,marginparwidth=0.5in]{geometry}

%\usepackage{lineno,hyperref}

\ifarchive
% Theorem definitions using amsthm
\usepackage{amsmath, amsthm, amsfonts}

@@

\else

\usepackage{amsmath, amsfonts}

\numberwithin{equation}{section}
\spnewtheorem{assump}{Assumption}{\bf}{\it}
\renewcommand{\theassump}{\Roman{assump}}
\spnewtheorem{prop}[equation]{Proposition}{\bf}{\rm}
\spnewtheorem{thrm}[equation]{Theorem}{\bf}{\it}
\spnewtheorem{defn}[equation]{Definition}{\bf}{\it}
\spnewtheorem{axiom}[equation]{Axiom}{\bf}{\it}
\spnewtheorem*{principle*}{Principle of scientific objectivity}{\bf}{\it}

\newenvironment{rationale}{\emph{Rationale}.}{\hfill\(\qed\)}
\newenvironment{justification}{\emph{Justification}.}{\hfill\(\qed\)}
\renewenvironment{proof}{\emph{Proof}.}{\hfill\(\qed\)}
\fi

\usepackage[only,llbracket, rrbracket,llparenthesis,rrparenthesis]{stmaryrd}

@@

% Allows to create negation symbols
\let\vec\undefined
\usepackage{MnSymbol}

@@

\title{Towards a general mathematical theory of experimental science}

\ifarchive
\author{Gabriele Carcassi, Christine A. Aidala \\ University of Michigan}
\date{\today}
\else
\author{Gabriele Carcassi \and Christine A. Aidala}
\institute{G. Carcassi, C. A. Aidala \at
	Physics Department, University of Michigan, Ann Arbor, MI 48109 \\
	\email{carcassi@umich.edu}
}
\date{Received: date / Accepted: date}
\fi

@@

\begin{abstract}
	We lay the groundwork for a formal theory that studies scientific theories, which, as physics provides the foundation for all sciences, we feel is a neglected responsibility of the physics community. We define a scientific theory as a set of verifiable statements, assertions that can be shown to be true with an experimental test in finite time. By studying the algebra of such objects, we show that verifiability already provides severe constraints. In particular, it requires that a set of physically distinguishable cases is naturally equipped with the mathematical structures (i.e. second countable Kolmogorov topologies and $\sigma$-algebras) that form the foundation of manifold theory, differential geometry, measure theory, probability theory and all the major branches of mathematics currently used in physics. This gives a clear physical meaning to those mathematical structures and provides a strong justification for their use in science. It also provides a formal framework to incorporate additional assumptions and constrain the search space for new physical theories.
\end{abstract}

Clarify our assumptions. As we are forced to formalize and make explicit our starting points, we understand under which conditions each physical theory is valid. We also have a better idea of when and how they fail, possibly giving new insights that may lead to new physics.
Put physics back at the center of the discussion. By starting from physical ideas we put physics back at the center of the discussion, instead of relegating it to a mere after-the-fact interpretation as it is currently done in most modern scientific theories. Many results can be directly understood in terms of the assumptions, fostering a better intuition for the mathematical results.
Give science sturdier mathematical grounds. As each mathematical symbol is given a precise physical meaning, our formal framework maps one-to-one to our physical understanding. This means we automatically leave out any unphysical mathematical artifact.
Foster connections between different fields of knowledge. We have found recurring core concepts between different fields of science, math, philosophy and engineering. In retrospect, this is not surprising: nature is one and does not care about such divisions, and therefore truly fundamental concepts must span freely across different areas.

@@

Stepping back and looking at physics as a whole, one can't fail to notice that it is essentially composed of different theories and models loosely connected to each other, each with its own starting points and realm of applicability. Classical mechanics, electromagnetism, general relativity, quantum mechanics, thermodynamics and so on may share some general ideas but, in the end, they all have their own separate foundation, either in a different set of laws and principles or simply in positing a unique mathematical structure. This state of affairs is so entrenched in our field that somehow feels like the best, if not the only, way to go. But is it? Or this approach actually prevents progress and understanding?

We can gain perspective comparing to different fields of science and engineering. For example, in computer science theory one starts by defining symbolic languages and computational devices~\cite{Turing} and then shows that no algorithm exists that can correctly decide whether an arbitrary program terminates given an arbitrary input~\cite{Sipser}. In communication theory one defines channels and the information they carry~\cite{Shannon} and proves general results such as calculating the maximum rate at which information can be transmitted over a channel given a set amount of noise~\cite{Pierce}. In control theory one defines a system in terms of state, inputs and outpust~\cite{Brogan} and then looks for general strategies for control such as Kalman filtering~\cite{Kalman}. These theories are \emph{general} in the sense that their results respectively apply to all control systems, all communication systems and all computational devices. Any topic in each field becomes a specialization of those general mathematical structure. In other words, the subject matter has been properly formally defined and the results always apply due to its mere definition.

In physics there is no such thing. Classical mechanics, thermodynamics and quantum mechanics are not the specialization of some more general theory. But how would that work? As in the other general theories, we need to clarify what our starting points are and let the logic follow. The first requirement of a physical theory, or rather of a scientific theory in general, is that it is experimentally testable. As mathematics is built on the logic of statements, we need to develop the logic of experimental verification. This would give the most general mathematical structure that serves as the baseline for everything else. The second requirement is having states and processes, which is where physics starts. We can then specialize these concepts for different cases. Some processes will be deterministic and reversible. Some processes will be non-deterministic and some of these will have equilibria. By adding different assumptions on states and process we would recover classical Hamiltonian mechanics in one case, quantum mechanics in another and thermodynamics in yet another. In all these theories, some basic properties will be common simply because we have states that need to be measured experimentally, and some properties will be different because the type of state or the type of processes are different. This is what we mean by a general mathematical theory of experimental science.

Such a general theory would provide a formal framework will force us to clarify our assumptions, clarifying the realm of applicability of each theory. It would physics back at the center of our discussion, as physical ideas are the starting point for our derivation and not after-the-fact interpretations. It would give science sturdier mathematical grounds, as each mathematical symbol is given a precise meaning and non-physical objects are excluded by construction. It would foster connections between different fields of knowledge: nature does not care about such divisions. It would provide a framework in which to pose new questions and solve old ones. In other words: it would provide a better foundations for physics and for the rest of the sciences. Developing a general mathematical theory for experimental science could be comparable to what happened in mathematics during the first half of the last century, when it was reorganized using logic and set theory as its foundation, which had profound repercussions in the field. Our preliminary work in both physics~\cite{Carc1} and math~\cite{Carc2} convinced us that such a goal is possible and within reach.

The aim of this work is to lay down the foundations for a general mathematical theory of experimental science. We start by defining verifiable statements: assertions whose truth can be tested experimentally in finite time. We study the logic of these statements, which is different from the standard Boolean logic because of termination. We define experimental domains as a collection of verifiable statement that, if given an indefinite amount of time, we can keep testing forever refining our knowledge. From each experimental domain we construct its possibilities: the statements that, if true, would give a prediction to all verifiable statements. These represents the different cases we can experimentally distinguish. The main results is that \textbf{each set of experimentally distinguishable objects comes equipped with a natural second countable Kolmogorov topology, where each open set corresponds to a verifiable statement, and a natural $\sigma$-algebra, where each Borel set corresponds to a theoretical statement which gives predictions to verifiable ones.} For example, \statement{the mass of the particle is between $0.4$ and $0.6$ MeV} is a verifiable statement precisely because $(0.4, 0.6)$ is an open set, while \statement{the mass of the particle is exactly $0.5$ MeV} is not verifiable because $[0.5,0.5]$ is not an open set.\statement{The standard topology of the real line keeps track of finite precision measurements} From that general result we can derive the following conclusion: \textbf{any set of physically distinguishable cases has most cardinality of continuum}. That is, the only thing we can do to distinguish a particular case is run a countable set of tests, whose output can be imagined as a countable sequence of true and false. This is equivalent to a real number from 0 to 1 expressed in base 2. In a nutshell, \textbf{experimental verification by itself guarantees us the existence of two mathematical structures that are the foundations of most tools used in physics}, such as differential geometry (Riemannian and symplectic), measure theory and probability theory and many other mathematical tools already used in physics and other sciences.

%In this paper we briefly present the starting point for our general mathematical theory of experimental science. It is the product of a decade spent in reverse engineering the laws of particle mechanics~\cite{Carc1} and our subsequent work on formalizing the starting points~\cite{Carc2}. Our overall program is large and ambitious and a journal article cannot treat it comprehensively. Therefore we will only present an overview of the basic concepts, which are already well consolidated and provide a good template to understand our approach. We will only provide a sketch for the proofs, which can be found in the draft of our open access book~\cite{Carc3} together with more details, discussions and insights.

While this result is remarkable by itself, it provides the basis for future expansions. Conceptually, in the same way that topological spaces keep track of what can be verified experimentally, manifolds keep track of objects that can be identified by continuous quantities, differentiable manifolds keep track of objects that allow a density to be defined over them and symplectic manifolds (i.e. phase space of classical mechanics) allow the density to be coordinate invariant. But by doing all this work formally we are forced to recognize the assumptions that go in those models. For example, we have already done preliminary work that uncovered a set of necessary and sufficient conditions such that a set of verifiable statements defines a continuous quantity. While, as one would expect, these can only correspond to an idealized case where one pretends a better precision measurement can always be achieved, they also tell us precisely how this idealized case fails which can be interesting for those working on plank scale physics. Specifically, ordering must fail which means not only that real number cannot be used to identify configurations in space and time, but that basic mathematical tools such as derivation and integration will break down as well. But, whatever other tools we will have, we \emph{know} we will always have a topology and a $\sigma$-algebra.

Another example, note that the set of discontinuous functions from $\mathbb{R}$ to $\mathbb{R}$ has cardinality greater than continuum and therefore cannot represent a set of physically distinguishable objects. This means that the square integrable space of functions typically used in quantum mechanics is far too big, and, in fact, it contains unphysical elements, such as states with infinite energy. However, the Schwartz space, like any set of continuous functions, has cardinality of continuum and can be given a second countable Kolmogorov topology.\footnote{It also has other properties that may map to physical requirements. For example, quantities of the form $\langle \psi|Q^n P^m|\psi\rangle$ will be always finite and the Fourier transform is well defined.} To our understanding, the reason why mathematically Hilbert spaces are useful is because one can take limits and use the vector space norm to converge. However, we should be able to use the corresponding $\sigma$-algebra to take limits, like one does in probability and measure theory. In other words, we may be able to construct, within this framework, a more mathematically sound and physically motivated foundation for quantum mechanics that includes the objects and only the objects that are physically meaningful. In turn, maybe this contribute to a solid mathematical foundation of quantum field theory which, we are told, is still an open problem.

As a last example, we should be able to form a connection to computer science: in the end a computation device is a physical system where the output is verifiable statement that is a function of the inputs, which also must be verifiable statements. In this context, all possible inputs form the possibilities of our space. Suppose we have a function from input to a Boolean. Then we can write the statements \statement{the input is such that the output of the function is true} and \statement{the input is such that the output of the function is false}. If both these statements correspond to open sets on the possible inputs, then we construct a test (i.e. a program) that will always terminate. That is, we can use all the tools in topology and computer science to answer a question of the type: is this problem testable or computable?

We hope that it is clear by now that a general mathematical theory of experimental science is missing, is possible and would be extremely useful. Whether you work on condensed matter, quantum thermodynamics, complex systems, particle physics, gravitation or grand unified theories, if what you are doing is science you will have, at some point, experimental verification. You will have problems (often the most interesting ones) that arise from questions such as: what is it that I can measure? How do I do it? Under what assumptions do the quantities in my theory map to those measurements? When are they even well defined? What can I compute within my theory? What is testable? Currently, you do not have a general framework to pose those questions. The idea that you will need a completely ad-hoc framework for your field only is not true. The mere \emph{logic} of experimental verification requires a particular structure. And there is a lot more to experimental verification than its logic: the quantity itself has to, at least in a sense, exist over a finite period of time to be measurable, there must be a process such that information about that quantity can be transferred to the measurement device, the system cannot be always assumed to be completely isolated as it would not be physically accessible. If you are working on foundational aspects, chances are you may already be thinking about some of these aspects as they relate to your field. Our question is: how much of those problems are specific to your field? How much can be treated generally, so that on one side each field has less work to do, and on the other side we have a common language across fields? What are the assumptions specific to your field and those that are more general? A general mathematical theory of experimental science would provide you with a formal framework to precisely pose and possibly answer those questions within your field. As the starting points are clarified, the foundations of different fields may be affected, or even partially integrated, and potentially change how they are taught.

The irony is that while the project is extremely ambitious, the biggest obstacles are not, at this point, technical. The biggest obstacles are sociological. On one side the nature of this work is very interdisciplinary. On the other side math and physics are divided in fields and sub-fields that are increasingly narrower.\footnote{We attended a topology conference where there seemed to be a disconnected between what mathematicians and condensed matter physicists meant by topology, to the point that they frequently talked past each other. It is also not uncommon to find scientists that work on gravitation that are not familiar with topological spaces, which are the foundation of differential geometry.} This mean that this work does not fit naturally in any community. One might think that this work belongs in mathematical physics, because we are working on the fundamental mathematical structures of physics. Yet, what most people actually mean by mathematical physics is partial differential equation and differential geometry. One might think that this work belongs in foundations of physics, because we are working on making explicit the fundamental assumptions one is taking. Yet, what most people actually mean by foundations of physics is interpretations of quantum mechanics, quantum gravity or some other ``quantum thingamajig''. This means that this type of work is most often rejected from journals or conferences without review because the editor feels the work is not in scope. Or that it is very difficult for journals to find reviewers with the right expertise. Or that no funding agency has a category for this type of work.\footnote{Curiously, work on theories that postulate parallel universes or on whether consciousness is what causes particles to decay is somehow well accepted in physics.} Or that arxiv.org categorizes this very paper as artificial intelligence without possibility of appeal. Or that, since noone has time and/or expertise to review on the merit, one merely looks at the celebrity of the authors. All of this becomes a self-fulfilling prophecy.

A common personal reaction from external observers, in fact, is to characterize this type of work as a new field of study. The problem is that normally new fields of study emerge from others as specialization or recombination. Therefore there is a time when physics gets discussed as part of natural philosophy, or quantum information is discussed as part of quantum mechanics or statistical mechanics. If sufficient critical mass is established, new fields, journal and funding categories are created. In our case, we are taking elements across too many different disciplines and putting them together to create something broader for each discipline to recognize itself in it. So we are most often told that our work \emph{is} interesting and worthwhile and it \emph{should} be done, but it should be done \emph{somewhere else} by \emph{someone else}. It is a strange form of Not In My Back Yard (NIMBY) that hits many who try to do interdisciplinary work and leaves us with no temporary home.

To obviate to these problems, borrowing practices from the open source software community, we are developing our main body of work in draft open access book form~\cite{Carc3}, so that we can address the topic in its entirety and our friends in philosophy, mathematics and physics can all check their parts. In fact, this very article is derived from that work and one should refer to it to find more mathematical and philosophical details. This type of peer review process is what is needed to get the job done, but it is impossible to do within traditional academic settings. It is only possible because we have found experts with very different backgrounds who value the work and understand its potential. For that we are lucky.

@@

	\bibitem{Carc2} C. A. Aidala, G. Carcassi, M. J. Greenfield: Topology and experimental distinguishability. Accepted by Topology Proceedings on 11 Dec 2018, arxiv:1708.05492.
